\documentclass{anstrans}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\title{Binary Formulation of Decay Equations}
\author{Anthony M. Scopatz$^{*}$, Cameron R. Bates$^{+}$, Paul P. H. Wilson$^{*}$}

\institute{
$^{*}$The University of Wisconsin-Madison, 1500 Engineering Drive, Madison,
WI\\
$^{+}$The University of California Berkeley, Etcheverry Hall, MC 1730, Berkeley, CA
}

\email{scopatz@wisc.edu}

%%%% packages and definitions (optional)
\usepackage{color}
\usepackage{graphicx} % allows inclusion of graphics
\usepackage{booktabs} % nice rules (thick lines) for tables
\usepackage{microtype} % improves typography for PDF
\usepackage{xspace}
\usepackage{listings}
\usepackage{textcomp}
\usepackage{ulem}

\usepackage[usenames,dvipsnames]{xcolor}


\definecolor{listinggray}{gray}{0.9}
\definecolor{lbcolor}{rgb}{0.9,0.9,0.9}
\lstset{
    %backgroundcolor=\color{lbcolor},
    language={Python},
    tabsize=4,
    rulecolor=\color{black},
    upquote=true,
    aboveskip={1.5\baselineskip},
    belowskip={1.5\baselineskip},
    columns=fixed,
    extendedchars=true,
    breaklines=true,
    prebreak=\raisebox{0ex}[0ex][0ex]{\ensuremath{\hookleftarrow}},
    frame=single,
    showtabs=false,
    showspaces=false,
    showstringspaces=false,
    basicstyle=\scriptsize\ttfamily\color{green!40!black},
    keywordstyle=\color[rgb]{0,0,1.0},
    commentstyle=\color[rgb]{0.133,0.545,0.133},
    stringstyle=\color[rgb]{0.627,0.126,0.941},
    numberstyle=\color[rgb]{0,1,0},
    identifierstyle=\color{black},
    captionpos=t,
}

\renewcommand{\vec}[1]{\bm{#1}} %vector is bold italic
\newcommand{\vd}{\bm{\cdot}} % slightly bold vector dot
\newcommand{\grad}{\vec{\nabla}} % gradient
\newcommand{\ud}{\mathop{}\!\mathrm{d}} % upright derivative symbol
\newcommand{\cyclus}{\textsc{Cyclus}\xspace}
\newcommand{\Cyclus}{\cyclus}

\begin{document}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Introduction}

The Bateman equations governing radioactive decay are an important 
subexpression of generalized transmutation equations. In many cases, it is 
desirable to compute decay on its own, outside of the presence of a neutron 
or photon field.  In this case radioactive decay is a function solely on 
intrinsic physical parameters, namely half-lives. This document recasts the 
Bateman equations into a form that is better suited for computation than the 
traditional expression.

This use case is particularly relevant to the study of the nuclear fuel 
cycle (NFC). While reactors are a critical component of most NFCs, there 
are many places in the cycle where fresh and used fuel is sitting outside 
of a core. In these cases, decay alone is needed. Full transmutation 
calculations require excessive computational effort when simply trying to 
model a storage facility. Thus, an efficient decay calculator has the 
potential to speed up simulation of the entire NFC.

Here, the traditional Bateman decay equations have been reformulated to 
an expression that is more suited for binary computation. Coupled with a 
novel code generation technique, this binary formulation provides 
dramatic speed increases in a modern computing environment. 

Notably, all of the software implemented for the decay calculation is 
available as part of the free \& open source PyNE library 
\cite{pyne2014,Bates2014}. Furthermore, this decay functionality has
been fully integrated into the \cyclus fuel cycle simulator
\cite{cyclus2015,cyclus_v1.2}.

This paper will start by reviewing the classic version of the Bateman 
equations in \S II before providing the desired departure from 
this form in \S III. Following this, implementation specific 
approximations are discussed in \S IV.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Canonical Bateman Equations for Decay}
\label{canon}
The canonical expression of the Bateman equations for a decay chain 
proceeding from a nuclide $A$ to a nuclide $Z$ at time 
$t$ following a specific path is as follows \cite{Cetnar2006640}:

\begin{equation}
\label{bm-eq}
N_C(t) = \frac{N_1(0)}{\lambda_C} \cdot \gamma \cdot \sum_{i=1}^C \lambda_i c_{i} e^{-\lambda_i t}
\end{equation}

\begin{table}[hbt]
\label{decay-symbol-meaning}
\caption{Symbol Meaning in Decay Equations}
\begin{tabular}{|l|l|}
\hline
\textbf{Symbol} & \textbf{Meaning} \\
\hline
$C$         & length of the decay chain\\
$i$         & index for ith species, on range [1, C]\\
$j$         & index for jth species, on range [1, C]\\
$t$         & time [seconds]\\
$N_i(t)$    & number density of the ith species at time t\\
$t_{1/2,i}$ & half-life of the ith species\\
$\lambda_i$ & decay constant of ith species, $ln(2)/t_{1/2,i}$\\
$\gamma$    & total branch ratio for this chain\\
\hline
\end{tabular}
\end{table}

The symbols in Equation \ref{bm-eq} have the meaning described in Table I.
Additionally, $c_{i}$ is defined as:
\begin{equation}
\label{c_i}
c_i = \prod_{j=1,i\ne j}^C \frac{\lambda_j}{\lambda_j - \lambda_i}
\end{equation}
Furthermore, the total chain branch ratio is defined as the product of the 
branch ratios between any two species in the chain \cite{harr2007precise}:

\begin{equation}
\label{gamma}
\gamma = \prod_{i=i}^{C-1} \gamma_{i \to i+1}
\end{equation}

Minor modifications are needed to Equation \ref{bm-eq} for terminal species: 
the first nuclide of a decay chain and the ending stable species. By setting 
$C=1$ for the first nuclide in a chain, the Bateman equations can be reduced 
to simply:%perhaps "the familiar" instead but it's also a bit colloquial

\begin{equation}
\label{N_C}
N_C(t) = N_1(0) e^{-\lambda_1 t}
\end{equation}

For stable species at the end of a chain, the appropriate equation is derived 
by taking the limit of when the decay constant of the stable nuclide 
($\lambda_C$) goes to zero.  Notice that in Equation \ref{c_i} that every $c_i$ contains exactly one $\lambda_C$
in the numerator which cancels with the $\lambda_C$ in the denominator 
in front of the summation in Equation \ref{bm-eq}. This leads to the following form 
for the calculation of the terminal species:

\begin{equation}
\label{lim_lam}
\lim_{\lambda_C \to 0} N_C(t) = N_1(0)  \gamma \left[e^{-0t} + \sum_{i=1}^{C-1} \lambda_i \left(\frac{1}{0 - \lambda_i} \prod_{j=1,i\ne j}^{C-1} \frac{\lambda_j}{\lambda_j - \lambda_i} \right) e^{-\lambda_i t} \right]
\end{equation}

\begin{equation}
\label{lim_lam_N_C}
N_C(t) = N_1(0)  \gamma \left[1.0 - \sum_{i=1}^{C-1} \left(\prod_{j=1,i\ne j}^{C-1} \frac{\lambda_j}{\lambda_j - \lambda_i} \right) e^{-\lambda_i t} \right]
\end{equation}

Thus, Equations \ref{bm-eq}, \ref{N_C}, \& \ref{lim_lam_N_C} together are
able to compute the concentration of any species under decay 
at any point in time.

\section{Binary Reformulation of Bateman Equations}
\label{bin}
There are two main strategies that may be used to construct a version of the 
Bateman equations that is better suited to computation, if not clarity. 

The first strategy is to minimize the number of arithmetic
operations that must be performed to achieve the same result. 
This can be done by grouping constants together and pre-calculating the 
result as its own derived constant. This saves the 
computer from having to perform these same arithmetic operations at run time.  
By grouping constants, it is possible to express the Bateman equations as a 
simple sum of exponentials, as seen in Equation \ref{N_C_bin}.

\begin{equation}
\label{N_C_bin}
N_C(t) = N_1(0) \sum_{i=1}^C k_{i} e^{-\lambda_i t}
\end{equation}

In Equation \ref{N_C_bin}, the coefficients $k_i$ are defined as follows:

\begin{equation}
\label{k_i}
k_i = \frac{\gamma}{\lambda_C} \lambda_i c_i
\end{equation}

If $k_i$ are computed at run time then this formulation results in much 
more computational effort that than the original Bateman equations. This is 
because the $\gamma/\lambda_C$ term is brought into the top-level summation 
in Equation \ref{N_C_bin}. This would increase the number of operations in the 
sum by $C$ to a total of $4C+1$, up from $3C+1$.
However, when the $k_i$s  are pre-calculated, 
the number of arithmetic operations for each term in the sum is a mere 4.
It is no longer a function of the length of the chain $C$ at all.  This is 
because explicitly computing $c_i$ and $\gamma$ have been completely avoided.
Note that even for the shortest non-stable species, this computation is 
equivalently expensive as the normal Bateman equations, namely 4 operations.
Thus even in the worst case, the binary formulation performs just as well.

The second strategy is to note that computers are much better at dealing 
with powers of 2 than any other base, even the natural base $e$. Thus the 
$\mathrm{exp2}(x)$ function, or $2^x$, is faster than the natural exponential 
function $\exp(x)$, $e^x$.  This is due to the number of floating point
operations performed in the implementation of the underlying exponential 
functions. As proof of this principle, the following simple timing results
demonstrate that $\mathrm{exp2}(x)$ is faster for random data:

\begin{lstlisting}[caption={Exponential Timing Comparison}, 
                   label=expcmp]
In [1]: import numpy as np

In [2]: r = np.random.random(1000) / np.random.random(1000)

In [3]: %timeit np.exp(r)
10000 loops, best of 3: 26.6 us per loop

In [4]: %timeit np.exp2(r)
10000 loops, best of 3: 20.1 us per loop
\end{lstlisting}

From here, $\mathrm{exp2}(x)$ enables a savings of about 25\% over 
$\exp(x)$. Note that this improvement is highly compiler and machine dependent.
However, since the core of the Bateman equations are exponentials, 
such a savings has the ability to reach all portions of a decay calculation.
Luckily, the decay constant itself provides an intrinsic mechanism to convert 
from base-e to base-2, as seen in the following equations:

\begin{equation}
\label{b2-0}
N_C(t) = N_1(0) \sum_{i=1}^C k_{i} \cdot e^{-\lambda_i t}
\end{equation}

\begin{equation}
\label{b2-1}
N_C(t) = N_1(0) \sum_{i=1}^C k_{i} \cdot \exp\left[\frac{-\ln(2)\cdot t}{t_{1/2,i}}\right]
\end{equation}

\begin{equation}
\label{b2-2}
N_C(t) = N_1(0) \sum_{i=1}^C k_{i} \cdot \mathrm{exp2}\left[\frac{-t}{t_{1/2,i}}\right]
\end{equation}

Equation \ref{b2-2} can be further collapsed by defining the constants $a$ to 
be the pre-computed exponent coefficient values:

\begin{equation}
\label{a_i}
a_i = \frac{-1}{t_{1/2,i}}
\end{equation}

Thus, the final form of the binary representation of the Bateman equations 
are shown in Equations \ref{nc_wakka}-\ref{nc_brown_shoes}:

\textbf{General Formulation:}

\begin{equation}
\label{nc_wakka}
N_C(t) = N_1(0) \sum_{i=1}^C k_{i} \cdot 2^{a_i t}
\end{equation}

\textbf{First Nuclide in Chain:}

\begin{equation}
\label{nc_jawakka}
N_C(t) = N_1(0) \cdot 2^{a_1 t}
\end{equation}

\textbf{Stable Nuclide:}

\begin{equation}
\label{nc_brown_shoes}
N_C(t) = N_1(0) \left[1.0 + \sum_{i=1}^{C-1} \lim_{\lambda_C\to 0}(k_{i}) \cdot 2^{a_i t} \right]
\end{equation}

With completely precomputed $k$, $a$, and the $\mathrm{exp2}()$ function, this 
formulation reduces the number of arithmetic operations to 3 for each term 
in the chain. Furthermore, it completely preserves all physical processes 
because no assumptions were made aside from the Bateman equations themselves.
Note that it is not possible to reduce the number of operations further
without assuming a value for $t$. Thus this representation is the 
mathematically fastest formulation of the Bateman equations.

\section{Implementation Specific Approximations}
\label{approx}
The method presented in the previous section has been implemented within
PyNE and Cyclus. However, any given mathematical system of equations
will encounter implementation-specific choices when turned into 
associated software. Thus, though the above formulation holds generally for 
any decay chain, certain approximations have been used in the implementation
for reasons of performance, compile-ability, and computability.
The PyNE implementation generally aims to reduce the number of 
chains and terms that are calculated when they can be shown to be 
redundant or insignificant to the total calculation. In specific, the 
following modeling assumptions have been made:

\begin{enumerate}
\item Decay chains coming from spontaneous fission are not tallied as they 
    lead to an explosion of the total number of chains while contributing to 
    extraordinarily rare branches.
\item Decay alphas are not treated as He-4 production.
\item For chains longer than length 2, any 
    term whose half-life is less than $10^{-8}$ of the sum of all 
    half-lives in the chain is dropped. This filtering prevents excessive
    calculation from species which do not significantly contribute to 
    end atom fractions. The relative threshold $10^{-8}$ was chosen
    because it is a reasonable naive estimate of floating point error after 
    many operations. If the filtering causes there to be less than 
    two terms in the summation, then the filtering is turned off and all
    terms are computed.
\item To prevent other sources of floating point error, a nuclide is 
    determined to be stable when $\lambda_i < 10^{-16}$, rather than when 
    $\lambda_i = 0.0$.
\item If a chain has any $\mathrm{NaN}$ decay constants, the chain is rejected.
\item If a chain has any infinite $k$ values, the chain is rejected.
\end{enumerate}

In principle, each of these statements is reasonable. However, they may 
have covariant effects that inject unreasonable errors into the system.
To ensure that this is not the case for the assumptions above, 
a benchmark study was performed \cite{benchmark}. This benchmark compared
the results of running ORIGEN v2.2 \cite{croff1980origen2} to the results of 
the binary formulation. This study compared 2781 cases for a wide spread of
nuclides over 0.01, 0.1, 1.0, 10, 100 times the half-life of the nuclide.
In the overwhelming majority of cases, the relative errors were found to be 
within acceptable bounds.  For cases that did prove significantly different, 
it differences could all be traced back to discrepancies in the underling 
data. Since the PyNE binary formulation uses more recent ENSDF decay 
data \cite{bhat1992evaluated} and computes more decay chains than ORIGEN, 
it is reasonable to assume that these discrepancies are not the fault of the 
binary formulation or the implementation assumptions.

It is also important to mention that these implementation assumptions 
may preclude desired behavior by users. For example, perhaps alpha decay 
should be considered He-4 production. This would bring the implementation in
line with the same feature in ORIGEN v2.2. In such a situation, these 
assumptions should be revisited and the implementation should be reworked.

\section{Comparison to ORIGEN implementation}

The implementation of decay calculations in ORIGEN follows a fundamentally 
different strategy from the one described above. It uses Taylor series expansion 
of the exponential decay to make the problem tractable
and then uses run-time branches to make a set of further approximations similar
to those described above. This is due to both the increased scope of the 
problem being solved, transmutation and decay, as well as the design which
treats the decay data and the algorithm as two separate entities. The 
combination of modern computational tools, allowing for efficient generation of 
code, and the existing PyNE decay data interface, with rapid access to up-to-date ENSDF
decay data, enable a merging of the two components and a reduction in computational
cost at run-time with a manageable computation cost at compile (less than 30 seconds on
a modern laptop). 
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Conclusions \& Future Work}

Many different computational solutions to the Bateman equations have been 
developed, but few have been focused on optimizing the solution for speed
of implementation without significant approximation of the initial problem.
 The ability to have algorithmically 
generated code makes it possible to generate a solution that is both 
computationally efficient and that minimizes mathematical approximations. This 
work demonstrates such an implementation and shows promising initial results
in comparison to existing tools. It is also unique in that the code requires no
run time access to nuclear data as it is all integrated into the compilation 
process. This results in a single portable C++ file with header that can be 
used in NFC tools for decay calculations. Future work will consist of a more 
detailed benchmark
comparison against existing tools and work to quantify the errors introduced by
the approximations used and those due to other effects such as round-off errors. 

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%\section{Acknowledgments}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\bibliographystyle{ans}
\bibliography{bibliography}
\end{document}
